A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-10-19 inst/manuscript/manuscript.tex

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Thereafter, several comparisons between
design-based and model-based for spatial data have been considered (Brus
and De Gruijter, 1997; Ver Hoef, 2002, 2008). Cooper (2006) review the
two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. In addition to Cooper (2006), there has been substantial
research and development into estimators that use both design and
model-based principles (see e.g.~Cicchitelli and Montanari (2012),
Chan-Golston et al. (2020) for a Bayesian approach, and Sterba (2009)).
More recent overviews include Brus (2020) and Wang et al. (2012).

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-akaike1974new}{}%
Akaike, H., 1974. A new look at the statistical model identification.
IEEE transactions on automatic control 19, 716--723.

@@

\leavevmode\hypertarget{ref-diggle2010geostatistical}{}%
Diggle, P.J., Menezes, R., Su, T.-l., 2010. Geostatistical inference
under preferential sampling. Journal of the Royal Statistical Society:
Series C (Applied Statistics) 59, 191--232.

>>>>

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Thereafter, several comparisons between
design-based and model-based for spatial data have been considered (Brus
and De Gruijter, 1997; Ver Hoef, 2002; Ver Hoef, 2008). Cooper (2006)
review the two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. In addition to Cooper (2006), there has been substantial
research and development into estimators that use both design and
model-based principles (see e.g. Cicchitelli and Montanari (2012),
Chan-Golston et al. (2020) for a Bayesian approach, and Sterba (2009)).
More recent overviews include Brus (2020) and Wang et al. (2012).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-akaike1974new}{}%
Akaike, H., 1974. A new look at the statistical model identification.
IEEE transactions on automatic control 19, 716--723.

@@

\leavevmode\hypertarget{ref-diggle2010geostatistical}{}%
Diggle, P.J., Menezes, R., Su, T., 2010. Geostatistical inference under
preferential sampling. Journal of the Royal Statistical Society: Series
C (Applied Statistics) 59, 191--232.
